% ======================================================================
% REFERENCES
% ======================================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Cheng \& Vlachos(2024)]{cheng2024entropy}
Cheng, J. \& Vlachos, A. (2024).
\newblock Measuring Uncertainty in Neural Machine Translation with Similarity-Sensitive Entropy.
\newblock In \emph{EACL 2024}, pp.\ 2115--2128.

\bibitem[Di~Gangi et al.(2019)]{digangi2019mustc}
Di~Gangi, M.A., Cattoni, R., Bentivogli, L., Negri, M., \& Turchi, M. (2019).
\newblock MuST-C: a Multilingual Speech Translation Corpus.
\newblock In \emph{NAACL 2019}, pp.\ 2012--2017.

\bibitem[Dou et al.(2024)]{dou2024iterative}
Dou, Q., Gales, J., et al. (2024).
\newblock Iterative Translation Refinement with Large Language Models.
\newblock In \emph{EAMT 2024}.

\bibitem[Fomicheva et al.(2020)]{fomicheva2020unsupervised}
Fomicheva, M., Sun, S., Yankovskaya, L., Blain, F., Guzm\'{a}n, F., Fishel, M., Aletras, N., Chaudhary, V., \& Specia, L. (2020).
\newblock Unsupervised Quality Estimation for Neural Machine Translation.
\newblock \emph{TACL}, 8, 539--555.

\bibitem[Ghazvininejad et al.(2019)]{ghazvininejad2019maskpredict}
Ghazvininejad, M., Levy, O., Liu, Y., \& Zettlemoyer, L. (2019).
\newblock Mask-Predict: Parallel Decoding of Conditional Masked Language Models.
\newblock In \emph{EMNLP-IJCNLP 2019}, pp.\ 6112--6121.

\bibitem[Gu et al.(2017)]{gu2017learning}
Gu, J., Neubig, G., Cho, K., \& Li, V.O.K. (2017).
\newblock Learning to Translate in Real-time with Neural Machine Translation.
\newblock In \emph{EACL 2017}, pp.\ 1053--1062.

\bibitem[Li et al.(2025)]{li2025tts_mt}
Li, Z., et al. (2025).
\newblock Test-Time Scaling of Reasoning Models for Machine Translation.
\newblock \emph{arXiv:2510.06471}.

\bibitem[Ma et al.(2019)]{ma2019stacl}
Ma, M., Huang, L., Xiong, H., Zheng, R., Liu, K., Zheng, B., Zhang, C., He, Z., Liu, H., Li, X., Wu, H., \& Wang, H. (2019).
\newblock STACL: Simultaneous Translation with Implicit Anticipation and Controllable Latency using Prefix-to-Prefix Framework.
\newblock In \emph{ACL 2019}, pp.\ 3025--3036.

\bibitem[Ma et al.(2020a)]{ma2020mma}
Ma, X., Pino, J., Cross, J., Puzon, L., \& Gu, J. (2020).
\newblock Monotonic Multihead Attention.
\newblock In \emph{ICLR 2020}.

\bibitem[Ma et al.(2020b)]{ma2020simulst}
Ma, X., Pino, J., \& Koehn, P. (2020).
\newblock SimulMT to SimulST: Adapting Simultaneous Text Translation to End-to-End Simultaneous Speech Translation.
\newblock In \emph{AACL-IJCNLP 2020}, pp.\ 582--587.

\bibitem[Ma et al.(2020c)]{ma2020simuleval}
Ma, X., Dousti, M.J., Wang, C., Gu, J., \& Pino, J. (2020).
\newblock SimulEval: An Evaluation Toolkit for Simultaneous Translation.
\newblock In \emph{EMNLP 2020 System Demonstrations}, pp.\ 144--150.

\bibitem[Madaan et al.(2023)]{madaan2023selfrefine}
Madaan, A., Tandon, N., Gupta, P., et al. (2023).
\newblock Self-Refine: Iterative Refinement with Self-Feedback.
\newblock In \emph{NeurIPS 2023}.

\bibitem[Papi et al.(2022)]{papi2022laal}
Papi, S., Gaido, M., Negri, M., \& Turchi, M. (2022).
\newblock Over-Generation Cannot Be Rewarded: Length-Adaptive Average Lagging for Simultaneous Speech Translation.
\newblock In \emph{AutoSimTrans 2022 (co-located with ACL 2022)}, pp.\ 12--17.

\bibitem[Papi et al.(2025)]{papi2025real}
Papi, S., Pol\'{a}k, P., Mach\'{a}\v{c}ek, D., \& Bojar, O. (2025).
\newblock How ``Real'' is Your Real-Time Simultaneous Speech-to-Text Translation System?
\newblock \emph{TACL}, 13, 281--313.

\bibitem[Rei et al.(2020)]{rei2020comet}
Rei, R., et al. (2020).
\newblock COMET: A Neural Framework for MT Evaluation.
\newblock In \emph{EMNLP 2020}, pp.\ 2685--2702.

\bibitem[Schuster et al.(2022)]{schuster2022calm}
Schuster, T., Fisch, A., Gupta, J., Dehghani, M., Bahri, D., Tran, V.Q., Tay, Y., \& Metzler, D. (2022).
\newblock Confident Adaptive Language Modeling.
\newblock In \emph{NeurIPS 2022}.

\bibitem[Snell et al.(2024)]{snell2024scaling}
Snell, C., Lee, J., Xu, K., \& Kumar, A. (2024).
\newblock Scaling LLM Test-Time Compute Optimally can be More Effective than Scaling Model Parameters.
\newblock \emph{arXiv:2408.03314; ICLR 2025}.

\bibitem[Wang et al.(2020)]{wang2020calibration}
Wang, S., Tu, Z., Shi, S., \& Liu, Y. (2020).
\newblock On the Inference Calibration of Neural Machine Translation.
\newblock In \emph{ACL 2020}, pp.\ 3070--3079.

\bibitem[Wang et al.(2021)]{wang2021covost2}
Wang, C., Wu, A., \& Pino, J. (2021).
\newblock CoVoST 2 and Massively Multilingual Speech-to-Text Translation.
\newblock In \emph{Interspeech 2021}, pp.\ 2247--2251.

\bibitem[Wang et al.(2023)]{wang2023selfconsistency}
Wang, X., Wei, J., Schuurmans, D., Le, Q., Chi, E., Narang, S., Chowdhery, A., \& Zhou, D. (2023).
\newblock Self-Consistency Improves Chain of Thought Reasoning in Language Models.
\newblock In \emph{ICLR 2023}.

\bibitem[Wei et al.(2022)]{wei2022cot}
Wei, J., Wang, X., Schuurmans, D., Bosma, M., Ichter, B., Xia, F., Chi, E., Le, Q., \& Zhou, D. (2022).
\newblock Chain-of-Thought Prompting Elicits Reasoning in Large Language Models.
\newblock In \emph{NeurIPS 2022}.

\bibitem[Xia et al.(2017)]{xia2017deliberation}
Xia, Y., Tian, F., Wu, L., Lin, J., Qin, T., Yu, N., \& Liu, T.-Y. (2017).
\newblock Deliberation Networks: Sequence Generation Beyond One-Pass Decoding.
\newblock In \emph{NeurIPS 2017}.

\end{thebibliography}